---
title: "Project title"
author:
  - name: Author Name
    affiliation:
    - &aff1 Author Institution, City, Country.
    email: email@website.com
date: "2026-02-23"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
vignette: >
  %\VignetteIndexEntry{Project title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Programming environment

```{r Set random seed, include=FALSE}
seed <- 2026-02-23
```

```{r Load packages, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggpubr)
library(pbapply)
library(readxl)
library(cluster)
library(broom)
library(brms)
library(cmdstanr) # Comment this if stand is not installable
library(broom.mixed) # Comment this if stand is not installable
library(igraph)
union <- base::union
library(ggnetwork)
```

```{r Load functions, include=FALSE}
lapply(list.files("R/", full.names = TRUE), source)
```

```{r Set theme, include=FALSE}
my_theme_set()
```

```{r Session information, echo=FALSE}
sessionInfo()
```

# Data Preprocessing

## Raw data

```{r Read code book, include=FALSE}
code_book <- read_csv("inst/extdata/code_book.csv", show_col_types = FALSE)
```

```{r Show code book, echo=FALSE}
show_table(code_book, "code_book", "Code book.")
```

```{r Column-type string, include=FALSE}
coltype_str <- paste0(code_book$col_type, collapse = "")
```

```{r Raw data, include=FALSE}
raw_data <- read_csv("inst/extdata/raw_data.csv", col_types = coltype_str)
```

```{r State row and column numbers of raw data, echo=FALSE}
cat("Rows: ", nrow(raw_data), "\n")
cat("Columns: ", ncol(raw_data))
```

## Select variable of interest

```{r Variable of interest, include=FALSE}
voi <- "pe"
```

```{r Rename the variable as outcome and put after identifier, include=FALSE}
proc_data <-
  raw_data |>
  rename(outcome = voi) |>
  select(id, outcome, everything())
```

# Data partition

```{r Determine partition sample size, include=FALSE}
whole_id <- proc_data$id
whole_n <- length(whole_id)
test_n <- ceiling(0.2 * whole_n)
dev_n <- whole_n - test_n
val_n <- ceiling(0.2 * dev_n)
train_n <- dev_n - val_n
```

```{r Randomly select identifiers for each partition, include=FALSE}
set.seed(seed)
test_id <- sample(whole_id, test_n)
dev_id <- setdiff(whole_id, test_id)

set.seed(seed)
val_id <- sample(dev_id, val_n)
train_id <- setdiff(dev_id, val_id)
```

```{r Create set data, include=FALSE}
set_data <-
  data.frame(id = train_id, set = "training") |>
  rbind(data.frame(id = val_id, set = "validation")) |>
  rbind(data.frame(id = test_id, set = "testing")) |>
  mutate(set = factor(set, c("training", "validation", "testing")))
```

```{r Join set data to outcome data, include=FALSE}
outcome_set_data <-
  proc_data |>
  select(id, outcome) |>
  left_join(set_data, by = "id")
```

```{r Summarize outcome class per set, include=FALSE}
if(class(outcome_set_data$outcome) == "numeric"){
  outcome_set_summary <-
    outcome_set_data |>
    group_by(set) |>
    summarize(n = n(), .groups = "drop")
  
  outcome_set_plot <-
    outcome_set_summary |>
    ggplot(aes(set, n)) +
    geom_col()
}else{
  outcome_set_summary <-
    outcome_set_data |>
    group_by(set, outcome) |>
    summarize(n = n(), .groups = "drop")
  
  outcome_set_plot <-
    outcome_set_summary |>
    ggplot(aes(set, n, fill = outcome)) +
    geom_col()
}
```

```{r Show summary per set, echo=FALSE, fig.height=90/25.4, fig.width=190/25.4}
outcome_set_plot
```

```{r Caption outcome class summary per set, echo=FALSE}
show_figure_legend("outcome_set_plot", "Data partition.")
```

```{r Split samples using the identifiers for each partition, include=FALSE}
dev_data <- slice(proc_data, match(dev_id, id))
train_data <- slice(proc_data, match(train_id, id))
val_data <- slice(proc_data, match(val_id, id))
test_data <- slice(proc_data, match(test_id, id))
```

# Data vizualization

```{r Select numerical variables, include=FALSE}
num_data <- select_if(train_data, is.numeric)
```

```{r Select categorical variables, include=FALSE}
cat_data <- select_if(train_data, is.factor)
```

## One-variable visualization

### QQ plot

```{r State number of QQ plots, echo=FALSE}
cat("Number of QQ plots (= # of numeric variables): ", ncol(num_data))
```

```{r Create custom function for QQ plot}
qq_plotter <- function(variable, data){
  data |>
    `colnames<-`(str_replace_all(colnames(data), variable, "variable")) |>
    ggplot(aes(sample = variable)) +
  	geom_qq() +
  	geom_qq_line() +
    xlab("Normal +/- SD") +
    ylab(variable) +
    theme(axis.title = element_text(size = 10))
}
```

```{r Create QQ plots, include=FALSE}
if(ncol(num_data) > 1){
  qq_plots <- pblapply(colnames(num_data), qq_plotter, num_data)
  
  qq_ncol <- 4
  qq_nrow <- ceiling(length(qq_plots) / qq_ncol)
  qq_plots <- ggarrange(plotlist = qq_plots, ncol = qq_ncol, nrow = qq_nrow)
}
```

```{r Show QQ plots, echo=FALSE, fig.height=270/25.4, fig.width=190/25.4}
if(ncol(num_data) > 1){
  qq_plots
}
```

```{r Caption QQ plots, echo=FALSE}
if(ncol(num_data) > 1){
  show_figure_legend("qq_plots", "QQ plots.")
}
```

### Bar chart

```{r State number of bar charts, echo=FALSE}
cat("Number of bar charts (= # of categorical variables): ", ncol(cat_data))
```

```{r Create custom function for bar chart}
bar_chart_plotter <- function(variable, data){
  data |>
    `colnames<-`(str_replace_all(colnames(data), variable, "variable")) |>
    group_by(variable) |>
    summarize(n = n()) |>
    mutate(p = n / sum(n) * 100) |>
    ggplot(aes(variable, p)) +
  	geom_col() +
  	xlab(variable) +
    ylab("Proportion (%)") +
    theme(axis.title = element_text(size = 10))
}
```

```{r Create bar charts, include=FALSE}
if(ncol(cat_data) > 1){
  bar_charts <- pblapply(colnames(cat_data), bar_chart_plotter, cat_data)
  
  bc_ncol <- 4
  bc_nrow <- ceiling(length(bar_charts) / bc_ncol)
  bar_charts <- ggarrange(plotlist = bar_charts, ncol = bc_ncol, nrow = bc_nrow)
}
```

```{r Show bar charts, echo=FALSE, fig.height=54/25.4, fig.width=190/25.4}
if(ncol(cat_data) > 1){
  bar_charts
}
```

```{r Caption bar charts, echo=FALSE}
if(ncol(cat_data) > 1){
  show_figure_legend("bar_charts", "Bar charts.")
}
```

## Two-to-three-variable visualization

### Scatter plot

```{r Select num vars for these hands-on, include=FALSE}
sel_num_var <- c("age", "bmi", "uta_ri", "uta_pi", "uta_psv", "sflt1", "plgf")
```

```{r Show selected num vars for these hands-on, echo=FALSE}
cat("Only these numeric variables are visualized:\n")
sel_num_var
```

```{r State number of scatter plots, echo=FALSE}
if(ncol(num_data) >= 2){
  combn_num_var <- combn(sel_num_var, 2, simplify = FALSE)
  combn_n <- length(combn_num_var)
}else{
  comb_n <- 0
}

cat("Number of scatter plots (= # of possible pairs of num. var.): ", combn_n)
```

```{r Create custom function for scatter plot}
scatter_plotter <- function(pair, data, third = NULL){
  if(is.null(third)){
    sc_plot <-
      data |>
      select_at(pair) |>
      `colnames<-`(c("V1", "V2")) |>
      ggplot(aes(V1, V2))
  }else{
    sc_plot <-
      data |>
      select_at(c(pair, third)) |>
      `colnames<-`(c("V1", "V2", "V3")) |>
      ggplot(aes(V1, V2, color = V3))
  }
  
  sc_plot <-
    sc_plot +
    geom_point(show.legend = FALSE)  +
  	xlab(pair[1]) +
    ylab(pair[2]) +
    theme(axis.title = element_text(size = 10))
  
  if(!is.null(third)){
    sc_plot <-
      sc_plot +
      scale_color_discrete(third)
  }
  
  return(sc_plot)
}
```

```{r Create scatter plots, include=FALSE}
if(ncol(num_data) >= 2){
  sc_plots <- pblapply(combn_num_var, scatter_plotter, num_data)
  
  sc_ncol <- 4
  sc_nrow <- ceiling(length(sc_plots) / sc_ncol)
  sc_plots <- ggarrange(plotlist = sc_plots, ncol = sc_ncol, nrow = sc_nrow)
}
```

```{r Show scatter plots, echo=FALSE, fig.height=270/25.4, fig.width=190/25.4}
if(ncol(num_data) >= 2){
  sc_plots
}
```

```{r Caption scatter plots, echo=FALSE}
if(ncol(num_data) >= 2){
  show_figure_legend("sc_plots", "Scatter plots.")
}
```

```{r Create scp with color, include=FALSE}
if(ncol(num_data) >= 2 & ncol(cat_data) >= 1){
  v3 <- colnames(cat_data)[1]
  v3_data <- select(cat_data, v3)
  num_v3_data <- cbind(v3_data, num_data)
  scc_plots <- pblapply(combn_num_var, scatter_plotter, num_v3_data, v3)
  
  scc_ncol <- 4
  scc_nrow <- ceiling(length(scc_plots) / scc_ncol)
  scc_plots <- ggarrange(plotlist = scc_plots, ncol = scc_ncol, nrow = scc_nrow)
}
```

```{r Show scp with color, echo=FALSE, fig.height=270/25.4, fig.width=190/25.4}
if(ncol(num_data) >= 2 & ncol(cat_data) >= 1){
  scc_plots
}
```

```{r Caption scp with color, echo=FALSE}
if(ncol(num_data) >= 2 & ncol(cat_data) >= 1){
  show_figure_legend("sc_plots", "Scatter plots.")
}
```

### Box and violin plots

```{r Select cat vars for these hands-on, include=FALSE}
sel_cat_var <- c("outcome", "iugr")
```

```{r Show selected cat vars for these hands-on, echo=FALSE}
cat("Only these categorical variables are visualized:\n")
sel_cat_var
```

```{r State number of box plots, echo=FALSE}
if(ncol(num_data) >= 1 & ncol(cat_data) >= 1){
  combn_num_cat_var <-
    expand.grid(V1 = sel_cat_var, V2 = sel_num_var) |>
    t() |>
    as.data.frame() |>
    as.list() |>
    `names<-`(NULL)
  
  combn_nc <- length(combn_num_cat_var)
}else{
  combn_nc <- 0
}

cat("Number of box plots (= # of possible pairs of num-cat. var.): ", combn_nc)
```

```{r Create custom function for box plot}
box_plotter <- function(pair, data){
  bx_plot <-
    data |>
    select_at(pair) |>
    `colnames<-`(c("V1", "V2")) |>
    ggplot(aes(V1, V2)) +
    geom_boxplot(outliers = FALSE)
  
  set.seed(seed)
  bx_plot <-
    bx_plot +
    geom_jitter(width = 0.2)
  
  bx_plot <-
    bx_plot +
  	xlab(pair[1]) +
    ylab(pair[2]) +
    theme(axis.title = element_text(size = 10))
  
  return(bx_plot)
}
```

```{r Create box plots, include=FALSE}
if(ncol(num_data) >= 1 & ncol(cat_data) >= 1){
  bx_plots <- pblapply(combn_num_cat_var, box_plotter, train_data)
  
  bx_ncol <- 4
  bx_nrow <- ceiling(length(bx_plots) / bx_ncol)
  bx_plots <- ggarrange(plotlist = bx_plots, ncol = bx_ncol, nrow = bx_nrow)
}
```

```{r Show box plots, echo=FALSE, fig.height=216/25.4, fig.width=190/25.4}
if(ncol(num_data) >= 1 & ncol(cat_data) >= 1){
  bx_plots
}
```

```{r Caption box plots, echo=FALSE}
if(ncol(num_data) >= 1 & ncol(cat_data) >= 1){
  show_figure_legend("bx_plots", "Box plots.")
}
```

```{r Create custom function for violin plot}
violin_plotter <- function(pair, data){
  vl_plot <-
    data |>
    select_at(pair) |>
    `colnames<-`(c("V1", "V2")) |>
    ggplot(aes(V1, V2)) +
    geom_violin()
  
  set.seed(seed)
  vl_plot <-
    vl_plot +
    geom_jitter(width = 0.2)
  
  vl_plot <-
    vl_plot +
  	xlab(pair[1]) +
    ylab(pair[2]) +
    theme(axis.title = element_text(size = 10))
  
  return(vl_plot)
}
```

```{r Create violin plots, include=FALSE}
if(ncol(num_data) >= 1 & ncol(cat_data) >= 1){
  vl_plots <- pblapply(combn_num_cat_var, violin_plotter, train_data)
  
  vl_ncol <- 4
  vl_nrow <- ceiling(length(vl_plots) / vl_ncol)
  vl_plots <- ggarrange(plotlist = vl_plots, ncol = vl_ncol, nrow = vl_nrow)
}
```

```{r Show violin plots, echo=FALSE, fig.height=216/25.4, fig.width=190/25.4}
if(ncol(num_data) >= 1 & ncol(cat_data) >= 1){
  vl_plots
}
```

```{r Caption violin plots, echo=FALSE}
if(ncol(num_data) >= 1 & ncol(cat_data) >= 1){
  show_figure_legend("vl_plots", "Violin plots.")
}
```

### Stacked bar chart

```{r State number of stacked bar charts, echo=FALSE}
if(ncol(cat_data) >= 2){
  combn_cat_var <- combn(sel_cat_var, 2, simplify = FALSE)
  combn_c <- length(combn_cat_var)
}else{
  combn_c <- 0
}

cat("Number of stacked bar charts (= # of poss. pairs of cat. var.): ", combn_c)
```

```{r Create custom function for stacked bar chart}
sb_chart_plotter <- function(pair, data, mode = "proportion"){
  
  data <-
    data |>
    select_at(pair) |>
    `colnames<-`(c("V1", "V2")) |>
    group_by(V1, V2) |>
    summarize(n = n(), .groups = "drop")
  
  if(mode == "count"){
    data <-
      data |>
      mutate(y = n)
    
    ylab_txt <- "Count"
  }else{
    data <-
      data |>
      group_by(V1) |>
      mutate(y = n / sum(n) * 100)
    
    ylab_txt <- "Proportion (%)"
  }
  
  data |>
    ggplot(aes(x = V1, y = y, fill = V2)) +
    geom_col() +
    xlab(pair[1]) +
    ylab(ylab_txt) +
    scale_fill_discrete(name = pair[2]) +
    theme(
      axis.title = element_text(size = 10)
      , legend.title = element_text(size = 10)
    )
}
```

```{r Create s-bar charts, include=FALSE}
if(ncol(cat_data) >= 2){
  sb_charts <- pblapply(combn_cat_var, sb_chart_plotter, cat_data)
  
  sb_ncol <- 3
  sb_nrow <- ceiling(length(sb_charts) / sb_ncol)
  sb_charts <- ggarrange(plotlist = sb_charts, ncol = sb_ncol, nrow = sb_nrow)
}
```

```{r Show s-bar charts, echo=FALSE, fig.height=54/25.4, fig.width=190/25.4}
if(ncol(cat_data) >= 2){
  sb_charts
}
```

```{r Caption s-bar charts, echo=FALSE}
if(ncol(cat_data) >= 2){
  show_figure_legend("sb_charts", "Stacked bar charts.")
}
```

```{r Create sc-bar charts, include=FALSE}
if(ncol(cat_data) >= 2){
  cb_charts <- pblapply(combn_cat_var, sb_chart_plotter, cat_data, "count")
  
  cb_ncol <- 3
  cb_nrow <- ceiling(length(cb_charts) / cb_ncol)
  cb_charts <- ggarrange(plotlist = cb_charts, ncol = cb_ncol, nrow = cb_nrow)
}
```

```{r Show sc-bar charts, echo=FALSE, fig.height=54/25.4, fig.width=190/25.4}
if(ncol(cat_data) >= 2){
  cb_charts
}
```

```{r Caption c-bar charts, echo=FALSE}
if(ncol(cat_data) >= 2){
  show_figure_legend("cb_charts", "Stacked bar charts by count.")
}
```

## More-than-three-variable visualization

```{r Create custom function to numerize categorical variables, include=FALSE}
numerize_cat_var <- function(x){
 as.numeric(x) - 1 
}
```

```{r Prepare cluster data, include=FALSE}
cluster_data <- train_data

if(ncol(cat_data) > 0){
  cluster_data <-
    cluster_data |>
    mutate_if(is.factor, numerize_cat_var) # Numerize categorical variables
}

cluster_data <-
  cluster_data |>
  column_to_rownames(var = "id") |>
  na.omit() # Remove rows with NA in any columns
```

## Sample clustering

```{r Prepare PC1 and PC2 for sample clustering, include=FALSE}
pca_fit <-
  cluster_data |>
  prcomp(scale. = TRUE) # Data standardization

pc12_data <-
  pca_fit$x |>
  as.data.frame() |>
  select(PC1, PC2)
```

```{r Use complete linkage for hierarchical clustering, include=FALSE}
hclust_fit <-
  pc12_data |>
  dist() |>
  hclust()
```

```{r Identify cluster labels, include=FALSE}
# Modify k accordingly
h_cluster <- cutree(hclust_fit, k = 3)
```

```{r Create custom function for PC-kmean plot}
pc_hclust_pca_plotter <- function(pc_data, h_cluster, pca_fit){
  
  var_exp <-
    summary(pca_fit)$importance["Proportion of Variance", 1:2] * 100
  
  pc_data |>
    mutate(cluster = factor(h_cluster)) |>
    ggplot(aes(x = PC1, y = PC2, color = cluster)) +
    geom_point(size = 3) +
    labs(
      x = paste0("PC1 (", round(var_exp[1], 1), "%)")
      , y = paste0("PC2 (", round(var_exp[2], 1), "%)")
      , color = "Hierarchical cluster"
    )
}
```

```{r Create PC-HC plot, include=FALSE}
pc_hclust_pca_plot <- pc_hclust_pca_plotter(pc12_data, h_cluster, pca_fit)
```

```{r Show PC-HC plot, echo=FALSE, fig.height=108/25.4, fig.width=190/25.4}
pc_hclust_pca_plot
```

```{r Caption PC-kmean plot, echo=FALSE}
show_figure_legend("pc_hclust_pca_plot", "PC-HC plot.")
```

```{r Create custom function for PC-hclust plot}
pc_hclust_plotter <- function(hclust_fit, h_cluster){
  h_cluster <- factor(h_cluster)
  cols <- as.integer(h_cluster)

  plot(hclust_fit, cex = 0.6)

  labels <- hclust_fit$labels
  order <- hclust_fit$order
  usr <- par("usr")

  y1 <- usr[3] - 0.02 * diff(usr[3:4])
  y2 <- usr[3] - 0.06 * diff(usr[3:4])

  text(
    x = seq_along(order)
    , y = y2
    , labels = h_cluster[order]
    , srt = 90
    , adj = 1
    , xpd = NA
    , cex = 0.6
    , col = cols[order]
  )
}
```

```{r Show PC-hclust plot, echo=FALSE, fig.height=190/25.4, fig.width=190/25.4}
pc_hclust_plotter(hclust_fit, h_cluster)
```

```{r Caption PC-hclust plot, echo=FALSE}
show_figure_legend("pc_hclust_plotter(hclust_fit)", "PC-hclust plot.")
```

# PCA and clustering analysis

```{r Create custom function to numerize categorical variables-2, include=FALSE}
numerize_cat_var <- function(x){
 as.numeric(x) - 1 
}
```

```{r Prepare cluster data-2, include=FALSE}
cluster_data <- train_data

if(ncol(cat_data) > 0){
  cluster_data <-
    cluster_data |>
    mutate_if(is.factor, numerize_cat_var) # Numerize categorical variables
}

cluster_data <-
  cluster_data |>
  column_to_rownames(var = "id") |>
  na.omit() # Remove rows with NA in any columns
```

## PCA

```{r Select variables for PCA}
# All variables except variable of interest
pca_var <- setdiff(colnames(cluster_data), "outcome")

# Selected variables (remove hash tag for commenting codes below)
# pca_var <- c("uta_ri", "uta_pi", "uta_pi", "sflt1", "plgf")
```

```{r Conduct PCA, include=FALSE}
pca_fit_var <-
  cluster_data |>
  select_at(pca_var) |>
  prcomp(scale. = TRUE) # Data standardization
```

```{r Create custom function to make scree plot}
scree_plotter <- function(pca){
  # Extract variance explained
  var_explained <-
    pca$sdev^2 / sum(pca$sdev^2)

  data <-
    data.frame(
      pc = seq_along(var_explained)
      , var = var_explained
    )

  ggplot(data, aes(x = pc, y = var)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = data$pc) +
    xlab("Principal Component") +
    ylab("Proportion of Variance Explained") +
    theme(
      axis.title = element_text(size = 10)
    )
}
```

```{r Show scree plot, echo=FALSE, fig.height=75/25.4, fig.width=190/25.4}
scree_plotter(pca_fit_var)
```

```{r Caption scree plot, echo=FALSE}
show_figure_legend("scree_plotter(pca_fit_var)", "Scree plot.")
```

```{r Create custom function to make score plot with optional grouping}
pc_score_plotter <- function(
  pca
  , pc_x = 1
  , pc_y = 2
  , group = NULL
){
  # Extract scores
  scores <-
    as.data.frame(pca$x)

  # Prepare data
  data <-
    scores |>
    mutate(
      group = if(is.null(group)) "All" else as.factor(group)
    )

  # Variance explained for labels
  var_explained <-
    pca$sdev^2 / sum(pca$sdev^2)

  xlab <-
    paste0(
      "PC", pc_x, " ("
      , round(var_explained[pc_x] * 100, 1)
      , "%)"
    )

  ylab <-
    paste0(
      "PC", pc_y, " ("
      , round(var_explained[pc_y] * 100, 1)
      , "%)"
    )

  ggplot(data, aes(
    x = .data[[paste0("PC", pc_x)]]
    , y = .data[[paste0("PC", pc_y)]]
    , color = group
  )) +
    geom_point(size = 3) +
    xlab(xlab) +
    ylab(ylab) +
    theme(
      axis.title = element_text(size = 10)
      , legend.title = element_blank()
    )
}
```

```{r Show score plot, echo=FALSE, fig.height=108/25.4, fig.width=190/25.4}
pc_score_plotter(pca_fit_var)
```

```{r Caption score plot, echo=FALSE}
show_figure_legend("pc_score_plotter(pca_fit_var)", "Score plot.")
```

```{r Create custom function to make loading plot}
pc_loading_plotter <- function(
  pca
  , pc_x = 1
  , pc_y = 2
  , labels = TRUE
){
  # Extract loadings
  loadings <-
    as.data.frame(pca$rotation)

  loadings$variable <-
    rownames(loadings)

  # Variance explained
  var_explained <-
    pca$sdev^2 / sum(pca$sdev^2)

  xlab <-
    paste0(
      "PC", pc_x, " ("
      , round(var_explained[pc_x] * 100, 1)
      , "%)"
    )

  ylab <-
    paste0(
      "PC", pc_y, " ("
      , round(var_explained[pc_y] * 100, 1)
      , "%)"
    )

  p <-
    ggplot(loadings) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +

    # Arrows
    geom_segment(
      aes(
        x = 0, y = 0
        , xend = .data[[paste0("PC", pc_x)]]
        , yend = .data[[paste0("PC", pc_y)]]
      )
      , arrow = arrow(length = unit(0.2, "cm"), type = "closed")
    ) +

    xlab(xlab) +
    ylab(ylab) +
    theme(
      axis.title = element_text(size = 10)
    )

  if(labels){
    p <-
      p +
      geom_text(
        aes(
          x = .data[[paste0("PC", pc_x)]]
          , y = .data[[paste0("PC", pc_y)]]
          , label = variable
        )
        , vjust = -0.5
        , size = 3
      )
  }

  return(p)
}
```

```{r Show loading plot, echo=FALSE, fig.height=190/25.4, fig.width=190/25.4}
pc_loading_plotter(pca_fit_var)
```

```{r Caption loading plot, echo=FALSE}
show_figure_legend("pc_loading_plotter(pca_fit_var)", "Loading plot.")
```

```{r Prepare PC1 and PC2 for clustering analysis, include=FALSE}
pc12_var_data <-
  pca_fit_var$x |>
  as.data.frame() |>
  select(PC1, PC2)
```

## Clustering analysis

```{r Choose between hierarhical and k-means clustering}
which_clustering <- c(`1` = "hierarchical", `2` = "k-means")[1]
```

```{r Choose clustering parameters}
# For hierarchical clustering
hclust_distance_method <-
  c(
    `1` = "gower"
    , `2` = "euclidean"
    , `3` = "binary"
    , `4` = "canberra"
    , `5` = "manhattan"
    , `6` = "maximum"
    , `7` = "minkowski"
  )[1]

hclust_agglomeration_method <-
  c(
    `1` = "average"
    , `2` = "ward.D2"
    , `3` = "complete"
    , `4` = "single"
  )[1]

h_cluster_n <- 4 # Change later after assessing the plot

# For k-means clustering
k_cluster <- NULL # If 2 clusters are expected, change to k_cluster <- 2

kmeans_k_method <-
  c(`1` = "elbow"
    , `2` = "silhouette"
  )[1]

kmeans_k_range <- 2:10

silhouette_distance_method <-
  c(
    `1` = "gower"
    , `2` = "euclidean"
    , `3` = "binary"
    , `4` = "canberra"
    , `5` = "manhattan"
    , `6` = "maximum"
    , `7` = "minkowski"
  )[1]
```

```{r Create a function to compute obtain WCSS}
kmeans_wcss <- \(k, data){
  set.seed(seed)
  fit <- kmeans(x = data, centers = k)
  
  data.frame(k = k, metric = fit$tot.withinss)
}
```

```{r Create a function to compute obtain silhouette score}
kmeans_silhouette_score <- \(k, data){
  set.seed(seed)
  fit <- kmeans(x = data, centers = k)
  
  d <-
    if(silhouette_distance_method == "gower"){
      cluster::daisy(data, metric = "gower")
    }else{
      dist(data, method = silhouette_distance_method)
    }

  sil <- silhouette(fit$cluster, d)

  data.frame(k = k, metric = mean(sil[, "sil_width"]))
}
```

```{r Create a function to plot elbow or silhouette method}
kmeans_k_plotter <- function(
  data
  , k_range = 2:10
  , method = c("elbow", "silhouette")
){
  method <- match.arg(method)

  if(method == "elbow"){
    metric_data <- map_dfr(k_range, \(k) kmeans_wcss(k, data))

    ylab <- "Total Within-Cluster Sum of Squares"
  }else{
    metric_data <- map_dfr(k_range, \(k) kmeans_silhouette_score(k, data))

    ylab <- "Average Silhouette Width"
  }

  ggplot(metric_data, aes(x = k, y = metric)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = k_range) +
    xlab("Number of Clusters (K)") +
    ylab(ylab) +
    theme(
      axis.title = element_text(size = 10)
    )
}
```

```{r Plot to choose k for k-means clustering if needed, include=FALSE}
if(which_clustering == "k-means" & is.null(k_cluster)){
  k_selection_plot <-
    kmeans_k_plotter(
      data = pc12_var_data
      , k_range = kmeans_k_range
      , method = kmeans_k_method
    )
}
```

```{r Show k plot, echo=FALSE, fig.height=90/25.4, fig.width=190/25.4}
if(which_clustering == "k-means" & is.null(k_cluster)){
  k_selection_plot
}
```

```{r Caption k plot, echo=FALSE}
if(which_clustering == "k-means" & is.null(k_cluster)){
  show_figure_legend("k_selection_plot", "k plot.")
}
```

```{r Conduct clustering analysis, include=FALSE}
if(which_clustering == "hierarchical"){
  if(hclust_distance_method == "gower"){
    clustering_fit <-
      pc12_data |>
      cluster::daisy(metric = "gower") |>
      hclust(method = hclust_agglomeration_method)
  }else{
    clustering_fit <-
      pc12_data |>
      dist(method = hclust_distance_method) |>
      hclust(method = hclust_agglomeration_method)
  }
}else if(which_clustering == "k-means"){
  set.seed(seed)
  clustering_fit <- kmeans(pc12_data, k_cluster)
}
```

```{r Obtain clusters, include=FALSE}
if(which_clustering == "hierarchical"){
  cluster_group <- cutree(clustering_fit, k = h_cluster_n)
}else{
  cluster_group <- clustering_fit$cluster
}
```

```{r Show PC-hcl plot EDA, echo=FALSE, fig.height=190/25.4, fig.width=190/25.4}
if(which_clustering == "hierarchical"){
  pc_hclust_plot <- pc_hclust_plotter(clustering_fit, cluster_group)
}
```

```{r Caption PC-hcl plot EDA, echo=FALSE}
if(which_clustering == "hierarchical"){
  show_figure_legend("pc_hclust_plot", "PC-hclust plot.")
}
```

```{r Create cl-score plot, include=FALSE}
cl_score_plot <- pc_score_plotter(pca_fit_var, group = cluster_group)
```

```{r Show cl-score plot, echo=FALSE, fig.height=108/25.4, fig.width=190/25.4}
cl_score_plot
```

```{r Caption cl-score plot, echo=FALSE}
show_figure_legend("cl_score_plot", "Clustered score plot.")
```

# Correlation matrix

## Normality assessment and test

### QQ plot

```{r Select numerical variables, include=FALSE}
num_data <- select_if(train_data, is.numeric)
```

```{r State number of QQ plots, echo=FALSE}
cat("Number of QQ plots (= # of numeric variables): ", ncol(num_data))
```

```{r Create custom function for QQ plot}
qq_plotter <- function(variable, data){
  data |>
    `colnames<-`(str_replace_all(colnames(data), variable, "variable")) |>
    ggplot(aes(sample = variable)) +
  	geom_qq() +
  	geom_qq_line() +
    xlab("Normal +/- SD") +
    ylab(variable) +
    theme(axis.title = element_text(size = 10))
}
```

```{r Create QQ plots, include=FALSE}
if(ncol(num_data) > 0){
  qq_plots <- pblapply(colnames(num_data), qq_plotter, num_data)
  
  qq_ncol <- 4
  qq_nrow <- ceiling(length(qq_plots) / qq_ncol)
  qq_plots <- ggarrange(plotlist = qq_plots, ncol = qq_ncol, nrow = qq_nrow)
}
```

```{r Show QQ plots, echo=FALSE, fig.height=270/25.4, fig.width=190/25.4}
if(ncol(num_data) > 0){
  qq_plots
}
```

```{r Caption QQ plots, echo=FALSE}
if(ncol(num_data) > 0){
  show_figure_legend("qq_plots", "QQ plots.")
}
```

### Data transformation

```{r Variables for squared-root test-transformation, include=FALSE}
ttrans_var_sqrt <- c("plgf", "sflt1_plgf")
```

```{r Variables for squared test-transformation, include=FALSE}
ttrans_var_sq <- c()
```

```{r Variables for log test-transformation, include=FALSE}
ttrans_var_log <- c()
```

```{r State test-ransformed variables, echo=FALSE}
ttrans_var <- c(ttrans_var_sqrt, ttrans_var_sq, ttrans_var_log)

cat(
  paste0("Test-transformed variables (", length(ttrans_var), "): ")
  , ttrans_var
)
```

```{r Create custom function for squared transformation}
sq <- function(x){
  x^2
}
```

```{r Data test-transformation using numerical train set}
if(ncol(num_data) > 0 & length(ttrans_var) > 0){
  num_data_ttrans <-
    num_data |>
    mutate_at(ttrans_var_sqrt, sqrt) |>
    mutate_at(ttrans_var_sq, sq) |>
    mutate_at(ttrans_var_log, log)
}
```

```{r Create QQ test-trans, include=FALSE}
if(ncol(num_data_ttrans) > 0 & length(ttrans_var) > 0){
  qt_plots <- pblapply(ttrans_var, qq_plotter, num_data_ttrans)
  
  qt_ncol <- 4
  qt_nrow <- ceiling(length(qt_plots) / qt_ncol)
  qt_plots <- ggarrange(plotlist = qt_plots, ncol = qt_ncol, nrow = qt_nrow)
}
```

```{r Show QQ test-trans, echo=FALSE, fig.height=54/25.4, fig.width=190/25.4}
if(ncol(num_data_ttrans) > 0 & length(ttrans_var) > 0){
  qt_plots
}
```

```{r Caption QQ test-trans, echo=FALSE}
if(ncol(num_data_ttrans) > 0 & length(ttrans_var) > 0){
  show_figure_legend("qt_plots", "QQ plots after data transformation.")
}
```

```{r Variables for squared-root transformation, include=FALSE}
trans_var_sqrt <- c("sflt1_plgf")
```

```{r Variables for squared transformation, include=FALSE}
trans_var_sq <- c()
```

```{r Variables for log transformation, include=FALSE}
trans_var_log <- c()
```

```{r State transformed variables, echo=FALSE}
trans_var <- c(trans_var_sqrt, trans_var_sq, trans_var_log)
cat(paste0("Transformed variables (", length(trans_var), "): "), trans_var)
```

```{r State categorized numerical variables, echo=FALSE}
cat_num_var <- setdiff(ttrans_var, trans_var)
cat_num_var <- c(cat_num_var, "uta_bn", "parity")
cat(
  paste0("Categorized numerical variables (", length(cat_num_var), "): ")
  , cat_num_var
)
```

```{r Create custom function categorize column using pre-defined cutoff}
categorize_from_table <- function(data, cutoff_table){
  
  if(nrow(cutoff_table) == 0) return(data)
  
  cols <- unique(cutoff_table$column)
  
  for(col_name in cols){
    
    rules <-
      cutoff_table |>
      filter(column == col_name) |>
      arrange(cutoff)
    
    breaks <- c(rules$cutoff, Inf)
    labels <- rules$category
    
    data[[col_name]] <-
      cut(
        data[[col_name]]
        , breaks = breaks
        , labels = labels
        , right = FALSE
        , include.lowest = TRUE
      )
  }
  
  mutate_at(data, cols, factor)
}
```

```{r Read pre-defined cutoff for categorization, include=FALSE}
category_cutoff <-
  read_csv("inst/extdata/category_cutoff.csv", show_col_types = FALSE)
```

```{r Transform data for all sets, include=FALSE}
if(ncol(num_data) > 0){
  
  proc_data_trans <- proc_data
  
  # Numerical data transformation
  if(length(trans_var) > 0){
    proc_data_trans <-
      proc_data_trans |>
      mutate_at(trans_var_sqrt, sqrt) |>
      mutate_at(trans_var_sq, sq) |>
      mutate_at(trans_var_log, log)
  }
  
  # Numerical data categorization
  code_book_trans <- code_book
  if(length(intersect(cat_num_var, unique(category_cutoff$column))) > 0){
    proc_data_trans <-
      proc_data_trans |>
      categorize_from_table(category_cutoff)
    
    code_book_trans <-
      read_csv("inst/extdata/code_book_trans.csv", show_col_types = FALSE)
  }
}
```

```{r Obtain train set from transformed dataset, include=FALSE}
train_data_trans <- slice(proc_data_trans, match(train_id, id))
```

### Determine Tukey's fences

```{r Select numerical variables from transformed dataset, include=FALSE}
num_data_trans <- select_if(train_data_trans, is.numeric)
```

```{r State number of numeric variables after transformation, echo=FALSE}
cat("Number of numeric variables after transformation: ", ncol(num_data_trans))
```

```{r Create custom function to determine Tukey fences}
tukey_fence <- function(x, which = c("lb", "ub"), na.rm = TRUE){
	which <- match.arg(which)

	q1 <- quantile(x, 0.25, na.rm = na.rm)
	q3 <- quantile(x, 0.75, na.rm = na.rm)
	iqr <- q3 - q1

	if(which == "lb"){
		q1 - 1.5 * iqr
	}else{
		q3 + 1.5 * iqr
	}
}
```

```{r Determine Tukey fences from training set, include=FALSE}
if(ncol(num_data_trans) > 0){
  tukey_fence_lb <-
    num_data_trans |>
    summarize_all(tukey_fence, which = "lb") |>
    gather(col_name, tukey_lb)
  
  tukey_fence_ub <-
    num_data_trans |>
    summarize_all(tukey_fence, which = "ub") |>
    gather(col_name, tukey_ub)
  
  tukey_fences <-
    tukey_fence_lb |>
    left_join(tukey_fence_ub, by = join_by(col_name))
}
```

### Outliers handling 

```{r Flag outliers using Tukey fences for all sets, include=FALSE}
outliers <-
  proc_data_trans |>
  select(colnames(num_data_trans)) |>
  mutate(seq = seq(n())) |>
  gather(col_name, value, -seq) |>
  left_join(tukey_fences, by = join_by(col_name)) |>
  mutate(is_outlier = value < tukey_lb | value > tukey_ub) |>
  select(seq, col_name, is_outlier) |>
  spread(col_name, is_outlier) |>
  arrange(seq) |>
  select(colnames(num_data_trans))
```

```{r Create dataset of outlier indicators, include=FALSE}
num_proc_data_outlier <-
  outliers |>
  mutate_all(as.numeric) |>
  mutate_all(as.factor) |>
  rename_all(paste0, "_ol")
```

```{r Obtain train set from outlier dataset, include=FALSE}
num_train_data_outlier <-
  select(proc_data_trans, id) |>
  cbind(num_proc_data_outlier) |>
  slice(match(train_id, id)) |>
  select(-id)
```

```{r Summarize outliers for train set, include=FALSE}
outliers_sum <-
  select(proc_data_trans, id) |>
  cbind(outliers) |>
  slice(match(train_id, id)) |>
  select(-id) |>
  summarize_all(mean, na.rm = TRUE) |>
  gather(variable, p) |>
  mutate(p = round(p * 100, 2))
```

```{r Show outliers summary, echo=FALSE}
show_table(outliers_sum, "outliers_sum", "Outliers summary in train set.")
```

```{r Remove outliers by assigning values to missing, include=FALSE}
num_proc_data_outlier_rm <-
  proc_data_trans |>
  select(colnames(num_data_trans))

num_proc_data_outlier_rm[as.matrix(outliers)] <- NA
```

```{r Bind back outlier-removed data for all sets, include=FALSE}
proc_data_outlier_rm <-
  proc_data_trans |>
  select(setdiff(colnames(proc_data_trans), colnames(num_data_trans))) |>
  cbind(num_proc_data_outlier_rm) |>
  select(colnames(proc_data_trans))
```

```{r Obtain train set from outlier-removed dataset, include=FALSE}
train_data_outlier_rm <- slice(proc_data_outlier_rm, match(train_id, id))
```

```{r Add outlier and missingness indicators, include=FALSE}
train_data_olms <-
  train_data_trans |>
  select_if(\(x) any(is.na(x))) |>
  mutate_all(is.na) |>
  mutate_all(as.numeric) |>
  mutate_all(factor) |>
  rename_all(paste0, "_ms")

train_data_olms <-
  train_data_outlier_rm |>
  cbind(num_train_data_outlier) |>
  cbind(train_data_olms)
```

### Assess normality by QQ plot and Shapiro-Wilk test

```{r Assess normality by QQ plot, include=FALSE}
normal_qq <- c("age", "height")
```

```{r State normal numerical variables by QQ plots, echo=FALSE}
cat("Normal numerical variables by QQ plot: ",normal_qq)
```

```{r List non-normal variables by QQ plot, include=FALSE}
non_normal_qq <- setdiff(colnames(num_data_trans), normal_qq)
```

```{r Conduct Shapiro-Wilk normality test, include=FALSE}
sw_results <-
  num_data_trans |>
  select_at(non_normal_qq) |>
  lapply(shapiro.test) |>
  lapply(tidy) |>
  imap(~ cbind(variable = .y, .x)) |>
  reduce(rbind) |>
  select(-method)
```

```{r Show Shapiro-Wilk normality test, echo=FALSE}
show_table(sw_results, "sw_results", "Shapiro-Wilk normality test results.")
```

```{r List normal variables by Shapiro-Wilk normality test, include=FALSE}
normal_sw <-
  sw_results |>
  filter(p.value > 0.05) |>
  pull(variable)
```

```{r List normal variables by Shapiro-Wilk normality test, include=FALSE}
non_normal_sw <- setdiff(non_normal_qq, normal_sw)
```

```{r Obtain normality for numerical columns, include=FALSE}
normality_data <-
  data.frame(col_name = colnames(num_data_trans)) |>
  mutate(normal = col_name %in% c(normal_qq, normal_sw))
```

## Perfect/partial separation

```{r Obtain scale for each column name except identifier, include=FALSE}
scale_data <-
    data.frame(
     scale =
       train_data_olms |>
       select(-id) |>
       sapply(class)
   ) |>
  rownames_to_column(var = "col_name")
```

```{r Identify all possible pair combinations, include=FALSE}
comb_pair <-
  combn(scale_data$col_name, 2) |>
  t() |>
  as.data.frame() |>
  left_join(
    rename(scale_data, V1 = col_name, scale_V1 = scale)
    , by = join_by(V1)
  ) |>
  left_join(
    rename(scale_data, V2 = col_name, scale_V2 = scale)
    , by = join_by(V2)
  )
```

```{r Create custom function to summarize variable pairs}
pair_summary <- function(data, V1, V2){
  result <-
    data |>
    group_by_at(c(V1, V2)) |>
    summarize(n = n(), .groups = "drop") |>
    `colnames<-`(c("value_V1", "value_V2", "n"))
  
  expand.grid(value_V1 = levels(data[[V1]]), value_V2 = levels(data[[V2]])) |>
    left_join(result, by = join_by(value_V1, value_V2)) |>
    mutate(n = ifelse(is.na(n), 0, n)) |>
    mutate(V1 = !!V1, V2 = !!V2)
}
```

```{r Perfect or partial separation data for categorical pairs, include=FALSE}
ps_data <-
  comb_pair |>
  filter(scale_V1 == "factor" & scale_V2 == "factor") |>
  select(V1, V2) |>
  pmap(\(V1, V2) pair_summary(train_data_olms, V1, V2)) |>
  reduce(rbind) |>
  group_by(V1, V2) |>
  summarize(ps = sum(n == 0) > 0, .groups = "drop")
```

## Correlation tests

```{r Obtain number of categories for categorical variables, include=FALSE}
level_data <-
  train_data_olms |>
  select_if(is.factor) |>
  summarize_all(\(x) length(intersect(levels(x), unique(x)))) |>
  gather(col_name, level_n)
```

```{r Identify variables with only 1 unique values, include=FALSE}
var_unique_1 <-
  train_data_olms |>
  select(-id) |>
  lapply(unique) |>
  lapply(setdiff, NA) |>
  sapply(length)

var_unique_1 <- var_unique_1[var_unique_1 < 2]
var_unique_1 <- names(var_unique_1)
```

```{r Identify var with 1 unique values after drop NA in pair, include=FALSE}
var_unique_1_na_pair0 <-
  comb_pair |>
  filter(!(V1 %in% var_unique_1 | V2 %in% var_unique_1))
  
var_unique_1_na_pair <-
  var_unique_1_na_pair0 |>
  select(V1, V2) |>
  pmap(\(V1, V2) select_at(train_data_olms, c(V1, V2))) |>
  lapply(unique) |>
  lapply(drop_na) |>
  lapply(lapply, unique) |>
  lapply(lapply, length) |>
  sapply(\(x) any(x < 2))

var_unique_1_na_pair <-
  var_unique_1_na_pair0 |>
  slice(which(var_unique_1_na_pair)) |>
  select(V1, V2)
```

```{r Identify mixed-scale pairs with only 1 sample in a cat, include=FALSE}
var_mixed_pair_1_cat_V10 <-
  comb_pair |>
  filter(scale_V1 == "factor" & scale_V2 == "numeric") |>
  select(V1, V2)

var_mixed_pair_1_cat_V1 <-
  var_mixed_pair_1_cat_V10 |>
  pmap(\(V1, V2) group_by_at(train_data_olms, c(V1, V2)))

var_mixed_pair_1_cat_V20 <-
  comb_pair |>
  filter(scale_V1 == "numeric" & scale_V2 == "factor") |>
  select(V1, V2)

var_mixed_pair_1_cat_V2 <-
  var_mixed_pair_1_cat_V20 |>
  pmap(\(V1, V2) group_by_at(train_data_olms, c(V2, V1)))

var_mixed_pair_1_cat <-
  list(V1 = var_mixed_pair_1_cat_V1, V2 = var_mixed_pair_1_cat_V2) |>
  lapply(lapply, summarize, n = n()) |>
  lapply(lapply, drop_na) |>
  lapply(lapply, `colnames<-`, c("cat", "num", "n")) |>
  lapply(lapply, group_by, cat) |>
  lapply(lapply, summarize, n = sum(n)) |>
  lapply(lapply, pull, n) |>
  lapply(sapply, \(x) any(x < 2))

var_mixed_pair_1_cat <-
  list(V1 = var_mixed_pair_1_cat_V10, V2 = var_mixed_pair_1_cat_V20) |>
  imap(~ slice(.x, which(var_mixed_pair_1_cat[[.y]]))) |>
  reduce(rbind)
```

```{r Identify untestable pairs, include=FALSE}
comb_pair_untestable <-
  comb_pair |>
  filter(
    V1 %in% var_unique_1 | V2 %in% var_unique_1
    | paste0(V1, V2) %in% unite(var_unique_1_na_pair, V, V1, V2, sep = "")$V
    | paste0(V1, V2) %in% unite(var_mixed_pair_1_cat, V, V1, V2, sep = "")$V
  )
```

```{r Join combinations with normality-separation-level data, include=FALSE}
comb_pair_normality_ps_level <-
  comb_pair |>
  anti_join(comb_pair_untestable, by = join_by(V1, V2, scale_V1, scale_V2)) |>
  left_join(
    rename(normality_data, V1 = col_name, normal_V1 = normal)
    , by = join_by(V1)
  ) |>
  left_join(
    rename(normality_data, V2 = col_name, normal_V2 = normal)
    , by = join_by(V2)
  ) |>
  left_join(ps_data, by = join_by(V1, V2))  |>
  left_join(
    rename(level_data, V1 = col_name, level_V1 = level_n)
    , by = join_by(V1)
  ) |>
  left_join(
    rename(level_data, V2 = col_name, level_V2 = level_n)
    , by = join_by(V2)
  ) |>
  mutate(
    normal_V1 = ifelse(is.na(normal_V1), TRUE, normal_V1)
    , normal_V2 = ifelse(is.na(normal_V2), TRUE, normal_V2)
    , ps = ifelse(is.na(ps), FALSE, ps)
    , level_V1 = ifelse(is.na(level_V1), 0, level_V1)
    , level_V2 = ifelse(is.na(level_V2), 0, level_V2)
  )
```

### Numerical-to-numerical pairs

```{r Identify numerical-to-numerical pairs, include=FALSE}
comb_pair_num_num <-
  comb_pair_normality_ps_level |>
  filter(scale_V1 == "numeric" & scale_V2 == "numeric")
```

```{r Identify normal-to-normal pairs, include=FALSE}
comb_pair_normal_normal <-
  comb_pair_num_num |>
  filter(normal_V1 & normal_V2) |>
  select(V1, V2)
```

```{r Identify non-normal-to-normal pairs, include=FALSE}
comb_pair_non_normal_normal <-
  comb_pair_num_num |>
  select(V1, V2) |>
  anti_join(comb_pair_normal_normal, join_by(V1, V2))
```

```{r Create custom function for correlation test}
correlation_test <- function(data, V1, V2, method){
  cor.test(data[[V1]], data[[V2]], method = "pearson") |>
    tidy() |>
    mutate(V1 = !!V1, V2 = !!V2) |>
    select(V1, V2, everything())
}
```

```{r Conduct Pearson correlation tests, include=FALSE}
if(nrow(comb_pair_normal_normal) > 0){
  cm_pearson <-
    comb_pair_normal_normal |>
    pmap(\(V1, V2) correlation_test(train_data_olms, V1, V2, "pearson")) |>
    reduce(rbind)
}
```

```{r Conduct Spearman correlation tests, include=FALSE}
if(nrow(comb_pair_non_normal_normal) > 0){
  cm_spearman <-
    comb_pair_non_normal_normal |>
    pmap(\(V1, V2) correlation_test(train_data_olms, V1, V2, "spearman")) |>
    reduce(rbind)
}
```

### Categorical-to-categorical pairs

```{r Identify categorical-to-categorical pairs, include=FALSE}
comb_pair_cat_cat <-
  comb_pair_normality_ps_level |>
  filter(scale_V1 == "factor" & scale_V2 == "factor")
```

```{r Identify pairs with separation, include=FALSE}
comb_pair_ps <-
  comb_pair_cat_cat |>
  filter(ps)
```

```{r Identify separation pairs with >2-category V1, include=FALSE}
comb_pair_ps_gt2_cat_V1 <-
  comb_pair_ps |>
  filter(level_V1 > 2) |>
  select(V1, V2)
```

```{r Create custom function for Bayesian categorical regression}
bayesian_cat_reg_fast <- function(data, V1, V2){
  data2 <-
    data |>
    select(all_of(c(V1, V2))) |>
    drop_na() |>
    droplevels()

  form <- reformulate(termlabels = V2, response = V1)

  result<-
    brm(
      formula = form
      , family = categorical()
      , data = data2
      , backend = "cmdstanr"
      , chains = 1
      , iter = 500
      , warmup = 250
      , refresh = 0
      , silent = 2
      , algorithm = "pathfinder"
    )
  
  tidy(result) |>
    mutate(V1 = !!V1, V2 = !!V2) |>
    select(V1, V2, everything())
}
```

```{r Conduct Bayesian categorical regression, include=FALSE}
if(nrow(comb_pair_ps_gt2_cat_V1) > 0){
  cm_bayesian_cat_reg <-
    comb_pair_ps_gt2_cat_V1 |>
    pmap(\(V1, V2) bayesian_cat_reg_fast(train_data_olms, V1, V2)) |>
    reduce(rbind)
}
```

```{r Identify separation pairs with 2-category V1, include=FALSE}
comb_pair_ps_2_cat_V1 <-
  comb_pair_ps |>
  filter(level_V1 == 2) |>
  select(V1, V2)
```

```{r Create custom function for Bayesian Bernoulli logistic regression}
bayesian_bern_reg_fast <- function(data, V1, V2){
  data2 <-
    data |>
    select(all_of(c(V1, V2))) |>
    drop_na() |>
    droplevels()

  form <- reformulate(termlabels = V2, response = V1)

  result <-
    brm(
      formula = form
      , family = bernoulli()
      , data = data2
      , backend = "cmdstanr"
      , chains = 1
      , iter = 500
      , warmup = 250
      , refresh = 0
      , silent = 2
      , algorithm = "pathfinder"
    )
  
  tidy(result) |>
    mutate(V1 = !!V1, V2 = !!V2) |>
    select(V1, V2, everything())
}
```

```{r Conduct Bayesian Bernoulli logistic regression, include=FALSE}
if(nrow(comb_pair_ps_2_cat_V1) > 0){
  cm_bayesian_bern_reg <-
    comb_pair_ps_2_cat_V1 |>
    pmap(\(V1, V2) bayesian_bern_reg_fast(train_data_olms, V1, V2)) |>
    reduce(rbind)
}
```

```{r Identify pairs without separation, include=FALSE}
comb_pair_non_ps <-
  comb_pair_cat_cat |>
  filter(!ps) |>
  select(V1, V2)
```

```{r Create custom function for Fisher exact test}
fisher_exact_test_fast <- function(data, V1, V2){
  result <-
    fisher.test(
      data[[V1]], data[[V2]]
      , workspace = 1e7, simulate.p.value = TRUE
    )
  
  tidy(result) |>
    select(p.value, method, alternative) |>
    mutate(V1 = !!V1, V2 = !!V2) |>
    select(V1, V2, everything())
}
```

```{r Conduct Fisher exact test, include=FALSE}
if(nrow(comb_pair_non_ps) > 0){
  cm_fisher <-
    comb_pair_non_ps |>
    pmap(\(V1, V2) fisher_exact_test_fast(train_data_olms, V1, V2)) |>
    reduce(rbind)
}
```

### Mixed-scale pairs

```{r Identify categorical-to-numerical pairs, include=FALSE}
comb_pair_cat_num <-
  comb_pair_normality_ps_level |>
  filter(scale_V1 == "factor" & scale_V2 == "numeric")
```

```{r Identify >2-categorical-to-norm-numerical pairs, include=FALSE}
comb_pair_cat_gt2_num_norm <-
  comb_pair_cat_num |>
  filter(level_V1 > 2) |>
  filter(normal_V2) |>
  select(V1, V2) |>
  mutate(inv = FALSE)
```

```{r Identify >2-categorical-to-non-norm-numerical pairs, include=FALSE}
comb_pair_cat_gt2_num_non_norm <-
  comb_pair_cat_num |>
  filter(level_V1 > 2) |>
  filter(!normal_V2) |>
  select(V1, V2) |>
  mutate(inv = FALSE)
```

```{r Identify 2-categorical-to-norm-numerical pairs, include=FALSE}
comb_pair_cat_2_num_norm <-
  comb_pair_cat_num |>
  filter(level_V1 == 2) |>
  filter(normal_V2) |>
  select(V1, V2) |>
  mutate(inv = FALSE)
```

```{r Identify 2-categorical-to-non-norm-numerical pairs, include=FALSE}
comb_pair_cat_2_num_non_norm <-
  comb_pair_cat_num |>
  filter(level_V1 == 2) |>
  filter(!normal_V2) |>
  select(V1, V2) |>
  mutate(inv = FALSE)
```

```{r Identify numerical-to-categorical pairs, include=FALSE}
comb_pair_num_cat <-
  comb_pair_normality_ps_level |>
  filter(scale_V1 == "numeric" & scale_V2 == "factor")
```

```{r Identify norm-numerical-to->2-categorical pairs, include=FALSE}
comb_pair_num_norm_cat_gt2 <-
  comb_pair_num_cat |>
  filter(level_V2 > 2)  |>
  filter(normal_V1) |>
  select(V1 = V2, V2 = V1) |>
  mutate(inv = TRUE)
```

```{r Identify non-norm-numerical-to->2-categorical pairs, include=FALSE}
comb_pair_num_non_norm_cat_gt2 <-
  comb_pair_num_cat |>
  filter(level_V2 > 2)  |>
  filter(!normal_V1) |>
  select(V1 = V2, V2 = V1) |>
  mutate(inv = TRUE)
```

```{r Identify norm-numerical-to-2-categorical pairs, include=FALSE}
comb_pair_num_norm_cat_2 <-
  comb_pair_num_cat |>
  filter(level_V2 == 2)  |>
  filter(normal_V1) |>
  select(V1 = V2, V2 = V1) |>
  mutate(inv = TRUE)
```

```{r Identify non-norm-numerical-to-2-categorical pairs, include=FALSE}
comb_pair_num_non_norm_cat_2 <-
  comb_pair_num_cat |>
  filter(level_V2 == 2)  |>
  filter(!normal_V1) |>
  select(V1 = V2, V2 = V1) |>
  mutate(inv = TRUE)
```

```{r Join >2-categorical and norm-numerical pairs, include=FALSE}
comb_pair_mixed_gt2_norm <-
  rbind(comb_pair_cat_gt2_num_norm, comb_pair_num_norm_cat_gt2)
```

```{r Create custom function for one-way ANOVA test}
one_way_anova_test <- function(data, V1, V2, inv){
  data <- data.frame(V1 = data[[V1]], V2 = data[[V2]])
  
  result <- aov(V2 ~ V1, data = data)
  
  result <-
    tidy(result) |>
    mutate(V1 = !!V1, V2 = !!V2)
  
  if(inv){
    result |>
      mutate(term = ifelse(term == "V1", "V2", term)) |>
      select(V1 = V2, V2 = V1, everything())
  }else{
    result |>
      select(V1, V2, everything())
  }
}
```

```{r Conduct one-way ANOVA test, include=FALSE}
if(nrow(comb_pair_mixed_gt2_norm) > 0){
  cm_one_way_anova <-
    comb_pair_mixed_gt2_norm |>
    pmap(\(V1, V2, inv) one_way_anova_test(train_data_olms, V1, V2, inv)) |>
    reduce(rbind)
}
```

```{r Join >2-categorical and non-norm-numerical pairs, include=FALSE}
comb_pair_mixed_gt2_non_norm <-
  rbind(comb_pair_cat_gt2_num_non_norm, comb_pair_num_non_norm_cat_gt2)
```

```{r Create custom function for Kruskal-Wallis test}
kruskal_wallis_test <- function(data, V1, V2, inv){
  V2_group_V1 <-
    data[[V1]] |>
    unique() |>
    lapply(\(x) data[[V2]][data[[V1]] == x])
  
  result <- kruskal.test(V2_group_V1)
  
  result <-
    tidy(result) |>
    mutate(V1 = !!V1, V2 = !!V2)
  
  if(inv){
    result |>
      select(V1 = V2, V2 = V1, everything())
  }else{
    result |>
      select(V1, V2, everything())
  }
}
```

```{r Conduct Kruskal-Wallis test, include=FALSE}
if(nrow(comb_pair_mixed_gt2_non_norm) > 0){
  cm_kruskal_wallis <-
    comb_pair_mixed_gt2_non_norm |>
    pmap(\(V1, V2, inv) kruskal_wallis_test(train_data_olms, V1, V2, inv)) |>
    reduce(rbind)
}
```

```{r Join 2-categorical and norm-numerical pairs, include=FALSE}
comb_pair_mixed_2_norm <-
  rbind(comb_pair_cat_2_num_norm, comb_pair_num_norm_cat_2)
```

```{r Create custom function for Welch two-sample t-test}
welch_two_sample_t_test <- function(data, V1, V2, inv){
  V2_group_V1_1 = data[[V2]][data[[V1]] == unique(data[[V1]])[1]]
  V2_group_V1_2 = data[[V2]][data[[V1]] == unique(data[[V1]])[2]]
  
  result <-
    try(
      t.test(
        V2_group_V1_1
        , V2_group_V1_2
        , alternative = "two.sided"
        , paired = FALSE
        , var.equal = FALSE
      )
    )
  
  if(is.character(result)){
    result <-
      data.frame(
        estimate = as.numeric(NA)
        , estimate1 = as.numeric(NA)
        , estimate2 = as.numeric(NA)
        , statistic = as.numeric(NA)
        , p.value = as.numeric(NA)
        , parameter = as.numeric(NA)
        , conf.low = as.numeric(NA)
        , conf.high = as.numeric(NA)
        , method =
          paste0(
            "Welch Two Sample t-test ("
            , paste0(result, collapse = ". ")
            , ")"
          )
        , alternative = "two.sided"
      )
  }else{
    result <- tidy(result)
  }
  
  result <- mutate(result, V1 = !!V1, V2 = !!V2)
  
  if(inv){
    result |>
      select(V1 = V2, V2 = V1, everything())
  }else{
    result |>
      select(V1, V2, everything())
  }
}
```

```{r Conduct Welch two-sample t-test, include=FALSE}
if(nrow(comb_pair_mixed_2_norm) > 0){
  cm_welch_two_sample_t <-
    comb_pair_mixed_2_norm |>
    pmap(\(V1, V2, inv) welch_two_sample_t_test(train_data_olms, V1, V2, inv)) |>
    reduce(rbind)
}
```

```{r Join 2-categorical and non-norm-numerical pairs, include=FALSE}
comb_pair_mixed_2_non_norm <-
  rbind(comb_pair_cat_2_num_non_norm, comb_pair_num_non_norm_cat_2)
```

```{r Create custom function for Wilcoxon rank-sum test}
wilcoxon_rank_sum_test <- function(data, V1, V2, inv){
  V2_group_V1_1 = data[[V2]][data[[V1]] == unique(data[[V1]])[1]]
  V2_group_V1_2 = data[[V2]][data[[V1]] == unique(data[[V1]])[2]]
  
  result <-
    try(
      wilcox.test(
        V2_group_V1_1
        , V2_group_V1_2
        , alternative = "two.sided"
        , paired = FALSE
      )
    )
  
  if(is.character(result)){
    result <-
      data.frame(
        statistic = as.numeric(NA)
        , p.value = as.numeric(NA)
        , method =
          paste0(
            "Welch Two Sample t-test ("
            , paste0(result, collapse = ". ")
            , ")"
          )
        , alternative = "two.sided"
      )
  }else{
    result <- tidy(result)
  }
  
  result <- mutate(result, V1 = !!V1, V2 = !!V2)
  
  if(inv){
    result |>
      select(V1 = V2, V2 = V1, everything())
  }else{
    result |>
      select(V1, V2, everything())
  }
}
```

```{r Conduct Welch two-sample t-test, include=FALSE}
if(nrow(comb_pair_mixed_2_non_norm) > 0){
  cm_wilcoxon_rank_sum <-
    comb_pair_mixed_2_non_norm |>
    pmap(\(V1, V2, inv) wilcoxon_rank_sum_test(train_data_olms, V1, V2, inv)) |>
    reduce(rbind)
}
```

### Multiple testing correction

```{r Create custom function to translate CI to p-value}
translate_ci_to_p_value <- function(data, V1, V2){
  data |>
    filter(V1 == !!V1 & V2 == !!V2) |>
    filter(!str_detect(term, "\\(Intercept\\)")) |>
    filter(!(conf.low <= 0 & conf.high >= 0)) |>
    summarize(n_sig = n()) |>
    mutate(p.value = ifelse(n_sig > 0, 0, 1)) |>
    filter(!is.na(p.value)) |>
    mutate(V1 = V1, V2 = V2) |>
    select(V1, V2, p.value)
}
```

```{r Translate CI to p-value for Bayesian cat regression, include=FALSE}
if(nrow(comb_pair_ps_gt2_cat_V1) > 0){
  cm_bayesian_cat_reg_p_value <-
    cm_bayesian_cat_reg |>
    select(V1, V2) |>
    unique() |>
    pmap(\(V1, V2) translate_ci_to_p_value(cm_bayesian_cat_reg, V1, V2)) |>
    reduce(rbind) |>
    mutate(method = "Bayesian categorical regression")
}
```

```{r Translate CI to p-value for Bayesian Bernoulli log reg, include=FALSE}
if(nrow(comb_pair_ps_2_cat_V1) > 0){
  cm_bayesian_bern_reg_p_value <-
    cm_bayesian_bern_reg |>
    select(V1, V2) |>
    unique() |>
    pmap(\(V1, V2) translate_ci_to_p_value(cm_bayesian_bern_reg, V1, V2)) |>
    reduce(rbind) |>
    mutate(method = "Bayesian Bernoulli logistic regression")
}
```

```{r Combine results from all correlation tests, include=FALSE}
correlation_results <- list()

if(nrow(comb_pair_untestable) > 0){
  correlation_results <-
    correlation_results |>
    c(list(
        untestable =
          comb_pair_untestable |>
          select(V1, V2) |>
          mutate(
            p.value = as.numeric(NA)
            , method =
              case_when(
                V1 %in% var_unique_1 | V2 %in% var_unique_1
                ~ paste0(
                    "None (untestable; "
                    , "1 variable only has 1 unique value after dropping "
                    , "missing values)"
                  )
                , paste0(V1, V2)
                  %in% unite(var_unique_1_na_pair, V, V1, V2, sep = "")$V
                  ~ paste(
                      "None (untestable; "
                      , "1 variable only has 1 unique value after dropping "
                      , "missing values in pair)"
                    )
                , paste0(V1, V2)
                  %in% unite(var_mixed_pair_1_cat, V, V1, V2, sep = "")$V
                  ~ paste0(
                      "None (untestable; "
                      , "a mixed-scale pairs with only 1 sample in 1 category"
                    )
                
              )
          )
      )
    )
}

if(!is.null(cm_pearson)){
  correlation_results <- c(correlation_results, list(pearson = cm_pearson))
}

if(!is.null(cm_spearman)){
  correlation_results <- c(correlation_results, list(spearman = cm_spearman))
}

if(nrow(comb_pair_ps_gt2_cat_V1) > 0){
  correlation_results <-
    c(correlation_results, list(bayesian_cat_reg = cm_bayesian_cat_reg_p_value))
}

if(nrow(comb_pair_ps_2_cat_V1) > 0){
  correlation_results <-
    correlation_results |>
    c(list(bayesian_bern_reg = cm_bayesian_bern_reg_p_value))
}

if(!is.null(cm_fisher)){
  correlation_results <- c(correlation_results, list(fisher = cm_fisher))
}

if(!is.null(cm_one_way_anova)){
  correlation_results <-
    correlation_results |>
    c(list(
        one_way_anova =
          cm_one_way_anova |>
          filter(term != "Residuals") |>
          mutate(method = "One-way ANOVA")
      )
    )
}

if(!is.null(cm_kruskal_wallis)){
  correlation_results <-
    c(correlation_results, list(kruskal_wallis = cm_kruskal_wallis))
}

if(!is.null(cm_welch_two_sample_t)){
  correlation_results <-
    c(correlation_results, list(welch_two_sample_t = cm_welch_two_sample_t))
}

if(!is.null(cm_wilcoxon_rank_sum)){
  correlation_results <-
    c(correlation_results, list(wilcoxon_rank_sum = cm_wilcoxon_rank_sum))
}
```

```{r Combine p-values from all correlation tests, include=FALSE}
correlation_p_values <-
  correlation_results |>
  lapply(select, V1, V2, p.value, method) |>
  reduce(rbind)
```

```{r Compute FDRs from p-values by Benjamini-Hochberg method, include=FALSE}
correlation_fdr <-
  correlation_p_values |>
  mutate(
    FDR =
      p.value |>
      p.adjust(method = "BH")
    , sig =
      ifelse(is.na(FDR), "Not tested", ifelse(FDR <= 0.05, "Yes", "No")) |>
      factor()
    , method_no =
      as.numeric(factor(method, unique(method))) |>
      factor()
  ) |>
  select(sig, FDR, everything())
```

### Visualize correlation by significance

```{r Create custom function to plot correlation matrix}
cor_matrix_plotter <- function(data, triangle = "lower"){
  vars <- union(as.character(data$V1), as.character(data$V2))

  data_plot <-
    data |>
    mutate(
      V1 = factor(V1, levels = vars),
      V2 = factor(V2, levels = rev(vars)),
      sig = factor(sig, levels = c("No", "Yes", "Not tested"))
    ) |>
    mutate(
      x_id = as.integer(factor(V1, levels = vars)),
      y_var = as.character(V2),
      y_id = match(y_var, rev(vars))
    )

  if(triangle == "lower"){
    data_plot <- data_plot |> filter(x_id + y_id <= length(vars) + 1)
  }else if(triangle == "upper"){
    data_plot <- data_plot |> filter(x_id + y_id >= length(vars) + 1)
  }

  ggplot(data_plot, aes(x = V1, y = V2, fill = sig)) +
    geom_tile(color = "white") +
    geom_text(aes(label = method_no), size = 2.5) +
    scale_fill_manual(
      values = c(
        "No" = "grey80",
        "Yes" = "steelblue",
        "Not tested" = "grey40"
      )
    ) +
    coord_equal() +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),
      axis.title = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    labs(fill = "Significant")
}
```

```{r Plot cor matrix, include=FALSE}
correlation_matrix <-
  comb_pair |>
  select(V1, V2) |>
  left_join(correlation_fdr, by = join_by(V1, V2)) |>
  cor_matrix_plotter()
```

```{r Show corr matrix, echo=FALSE, fig.height=150/25.4, fig.width=190/25.4}
correlation_matrix
```

```{r Caption corr matrix, echo=FALSE}
show_figure_legend("correlation_matrix", "Correlation matrix.")
```

```{r Obtain correlation method numbers, include=FALSE}
cm_method_no <-
  correlation_matrix$data |>
  select(no = method_no, method) |>
  unique() |>
  arrange(no)
```

```{r Show correlation method numbers, echo=FALSE}
show_table(cm_method_no, "cm_method_no", "Correlation methods.")
```

## Hypothesis generation from significant correlations

```{r Filter significant correlations, include=FALSE}
correlations <-
  correlation_matrix$data |>
  filter(sig == "Yes") |>
  filter(!(str_detect(V1, "_ol$|ms$") | str_detect(V2, "_ol$|_ms$"))) |>
  select(V1, V2)
```

```{r Create custom function to plot correlation network}
correlation_network_plotter <- function(
  correlations
  , layout = "fr"
  , focal = NULL   # length 2
  , directed = FALSE
  , title = NULL
  , curvature = 0
){
  nodes <-
    tibble(
      name = sort(unique(c(correlations$V1, correlations$V2)))
    )

  edges <-
    correlations |>
    transmute(
      from = V1,
      to = V2
    ) |>
    distinct()

  graph <-
    graph_from_data_frame(
      d = edges,
      vertices = nodes,
      directed = FALSE
    )

  set.seed(seed)

  if(layout == "fr"){
    layout_xy <-
      layout_with_fr(graph) |>
      as.data.frame()

  }else if(layout == "tree"){

    if(is.null(focal) || length(focal) != 2){
      stop("For tree layout, provide focal = c(var1, var2)")
    }

    # ensure focals exist
    if(!all(focal %in% nodes$name)){
      stop("Focal variables not found in nodes")
    }

    # build BFS order starting from both focals
    root_ids <- match(focal, nodes$name)

    bfs_order <- unique(unlist(
      lapply(root_ids, function(r){
        bfs(graph, root = r, order = TRUE)$order
      })
    ))

    layout_xy <-
      layout_as_tree(
        graph
        , root = root_ids
      ) |>
      as.data.frame()

  }else{
    stop("layout must be 'fr' or 'tree'")
  }

  colnames(layout_xy) <- c("x", "y")

  node_data <-
    nodes |>
    mutate(
      x = layout_xy$x,
      y = layout_xy$y
    )
  
  edge_data <-
    edges |>
    left_join(
      rename(node_data, x = x, y = y, from = name),
      by = join_by(from)
    ) |>
    left_join(
      rename(node_data, xend = x, yend = y, to = name),
      by = join_by(to)
    )
  
  if(!is.null(focal)){
    node_data <-
      node_data |>
      mutate(
        voi =
          ifelse(
            name %in% focal
            , paste0("Focal ", as.numeric(factor(name, focal)))
            , "Non-focal"
          )
      )
    
    edge_data <-
      edge_data |>
      mutate(
        voi =
          ifelse(
            from %in% focal
            , paste0("Focal ", as.numeric(factor(from, focal)))
            , "Non-focal"
          )
      )
  }
  
  edge_data <-
    edge_data |>
    filter(!(abs(x - xend) < 1e-12 & abs(y - yend) < 1e-12))
  
  p <- ggplot()
  
  if(!is.null(focal)){
    p <-
      p +
      geom_nodes(data = node_data, aes(x, y, color = voi), size = 6)
  }else{
    p <-
      p +
      geom_nodes(data = node_data, aes(x, y), size = 6)
  }
    
  
  if(directed){
    p <-
      p +
      geom_edges(
        data = edge_data,
        aes(x, y, xend = xend, yend = yend, color = voi),
        arrow = arrow(length = unit(6, "pt"), type = "closed"),
        curvature = curvature
      )
  }else{
    p <-
      p +
      geom_edges(
        data = edge_data,
        aes(x, y, xend = xend, yend = yend),
        curvature = curvature
      )
  }
  
  if(!is.null(focal)){
    p <-
      p +
      geom_label(
        data = node_data, aes(x, y, label = name, color = voi),
        size = 3, hjust = -0.2, vjust = 1.2, show.legend = FALSE
      )
  }else{
    p <-
      p +
      geom_label(
        data = node_data, aes(x, y, label = name),
        size = 3, hjust = -0.2, vjust = 1.2, show.legend = FALSE
      )
  }
  
  p +
    scale_color_discrete("Variable of\ninterest") +
    coord_cartesian(clip = "off") +
    theme_blank() +
    ggtitle(title)
}
```

```{r Plot cor net for VOI, include=FALSE}
correlation_network_voi <-
  correlations |>
  filter(V1 == "outcome" | V2 == "outcome") |>
  correlation_network_plotter()
```

```{r Show cor net for VOI, echo=FALSE, fig.height=90/25.4, fig.width=190/25.4}
correlation_network_voi
```

```{r Caption cor net for VOI, echo=FALSE}
show_figure_legend("correlation_network_voi", "Correlation network for VOI without showing edges among non-VOI variables.")
```

```{r Variable 2 of interest for correlation of interest (COI), include=FALSE}
voi2 <- "plgf"
```

```{r Determine correlation of interest (COI), include=FALSE}
coi <-
  correlations |>
  filter(V1 == "outcome"| V2 == "outcome" | V1 == voi2 | V2 == voi2)
```

```{r Plot cor net for COI, include=FALSE}
correlation_network_coi <-
  correlation_network_plotter(coi, focal = c("outcome", voi2))
```

```{r Show cor net for COI, echo=FALSE, fig.height=90/25.4, fig.width=190/25.4}
correlation_network_coi
```

```{r Caption cor net for COI, echo=FALSE}
show_figure_legend("correlation_network_coi", "Correlation network for COI related to two VOIs without showing edges among non-VOI variables.")
```

```{r Identify statistical confounders, include=FALSE}
stat_confounders <-
  list(
    filter(coi, V1 == "outcome"| V2 == "outcome")
    , filter(coi, V1 == voi2 | V2 == voi2)
  ) |>
  lapply(mutate_all, as.character) |>
  lapply(lapply, unique) |>
  lapply(reduce, c) |>
  lapply(setdiff, c("outcome", voi2)) |>
  reduce(intersect)
```

```{r Determine VOIs with confounders (VCF), include=FALSE}
vcf <-
  coi |>
  filter(
    V1 %in% c("outcome", voi2, stat_confounders)
    & V2 %in% c("outcome", voi2, stat_confounders)
  )
```

```{r Plot cor net for VCF, include=FALSE}
correlation_network_vcf <-
  correlation_network_plotter(vcf, "tree", c("outcome", voi2))
```

```{r Show cor net for VCF, echo=FALSE, fig.height=50/25.4, fig.width=190/25.4}
correlation_network_vcf
```

```{r Caption cor net for VCF, echo=FALSE}
show_figure_legend("correlation_network_vcf", "Correlation network for VCF related to two VOIs without showing edges among non-VOI variables.")
```

# Causal graph

## Causal, mediator, confounder, and collider paths

```{r Create undirected VOIs with confounders, include=FALSE}
vcf_undirected <- mutate(vcf, from = NA)
```

```{r Write the undirected VOIs to CSV, include=FALSE}
write_csv(vcf_undirected, "inst/extdata/vcf_undirected.csv")
```

```{r Determine causal direction per pair related to VOIs, include=FALSE}
vcf_directed <-
  read_csv("inst/extdata/vcf_directed.csv", show_col_types = FALSE)
```

```{r Obtain causal diagram data from VCF, include=FALSE}
causal_diagram_vcf_data <-
  vcf_directed |>
  mutate(to = ifelse(from == 1, 2, 1)) |>
  mutate(
    from = ifelse(from == 1, V1, V2)
    , to = ifelse(to == 1, V1, V2)
  ) |>
  select(V1 = from, V2 = to)
```

```{r Plot causal diagram for VCF, include=FALSE}
causal_diagram_vcf_plot <-
  causal_diagram_vcf_data |>
  correlation_network_plotter(
    "tree"
    , focal = c("outcome", voi2)
    , directed = TRUE
  )
```

```{r CD for VCF, echo=FALSE, fig.height=50/25.4, fig.width=190/25.4}
causal_diagram_vcf_plot
```

```{r Caption CD for VCF, echo=FALSE}
show_figure_legend("causal_diagram_vcf_plot", "Causal diagram for VCF related to two VOIs without showing edges among non-VOI variables.")
```

```{r Determine causal-mediator-confounder-collider of 2 VOIs, include=FALSE}
path <-
  read_csv("inst/extdata/path.csv", show_col_types = FALSE) |>
  group_by(path) |>
  summarise(values = list(variable)) |>
  deframe()
```

## Causation related to measurement variables

```{r Create a function to connect two outlier or missingness indicators}
olms_connection <- function(data, suffix){
  data |>
    mutate_all(as.character) |>
    mutate(
      from =
        case_when(
          str_detect(V1, suffix) & str_detect(V2, suffix) ~ paste0(V1, "|", V2)
          , str_detect(V1, suffix) & !str_detect(V2, suffix) ~ V2
          , TRUE ~ V1
        )
      , to =
        case_when(
          str_detect(V2, suffix) & str_detect(V1, suffix) ~ paste0(V2, "|", V1)
          , str_detect(V2, suffix) & !str_detect(V1, suffix) ~ V2
          , TRUE ~ V1
        )
    ) |>
    separate_rows(from, sep = "\\|") |>
    separate_rows(to, sep = "\\|") |>
    filter(from != to) |>
    select(V1 = from, V2 = to) |>
    unique()
}
```

### Measurement error

```{r Filter significant correlations of outlier indicators, include=FALSE}
vcf_ol <- paste0(unique(c(vcf$V1, vcf$V2)), "_ol")

correlations_ol <-
  correlation_matrix$data |>
  filter(sig == "Yes") |>
  filter(V1 %in% vcf_ol | V2 %in% vcf_ol) |>
  select(V1, V2)
```

```{r Determine causal direction with outlier indicators, include=FALSE}
ol_directed <- olms_connection(correlations_ol, "_ol$")
```

```{r Determine ol-causal direction with measurement variables, include=FALSE}
o_directed <-
  data.frame(V1 = unique(ol_directed$V2)) |>
  mutate(V2 = str_replace_all(V1, "_ol$", "_m"))

o_directed <-
  data.frame(
    V1 = unique(c(causal_diagram_vcf_data$V1, causal_diagram_vcf_data$V2))
  ) |>
  mutate(V2 = paste0(V1, "_m")) |>
  rbind(o_directed)
```

```{r Determine ol-causal direction of all with outliers, include=FALSE}
o_directed_all <-
  data.frame(V1 = colnames(train_data_olms)) |>
  filter(str_detect(V1, "_ol$")) |>
  mutate(V2 = str_replace_all(V1, "_ol$", "_m"))

o_directed_all <-
  o_directed_all |>
  mutate(V1 = str_remove_all(V1, "_ol$")) |>
  rbind(o_directed_all)
```

```{r Obtain CD data with outlier indicators, include=FALSE}
causal_diagram_ol_data <-
  causal_diagram_vcf_data |>
  rbind(o_directed) |>
  rbind(ol_directed) |>
  rbind(o_directed_all) |>
  unique()
```

```{r Create custom function to assess outlier by CD}
assess_ol <- function(data, focal, title = NULL){
  focal <- paste0(focal, c("", "_ol", "_m"))
  
  data <-
    data |>
    filter(V1 %in% focal | V2 %in% focal)
  
  if(nrow(data) == 0){
    NULL
  }else{
    data |>
      correlation_network_plotter(
        layout = "tree"
        , focal = focal[2:3]
        , directed = TRUE
        , title = title
        , curvature = 0.1
      )
  }
}
```

```{r Plot CD with outlier indicators, include=FALSE}
causal_diagram_ol_plot <-
  o_directed_all |>
  filter(str_detect(V1, "_ol$")) |>
  pull(V1) |>
  str_remove_all("_ol$")

causal_diagram_ol_plot <-
  causal_diagram_ol_plot |>
  `names<-`(causal_diagram_ol_plot)

causal_diagram_ol_plot <-
  causal_diagram_ol_plot |>
  imap(~ assess_ol(causal_diagram_ol_data, .x, .y))

causal_diagram_ol_plot <-
  causal_diagram_ol_plot[!sapply(causal_diagram_ol_plot, is.null)]

causal_diagram_ol_plot <-
  ggarrange(
    plotlist = causal_diagram_ol_plot
    , nrow = length(causal_diagram_ol_plot)
  )
```

```{r CD with OL, echo=FALSE, fig.height=850/25.4, fig.width=190/25.4}
causal_diagram_ol_plot
```

```{r Determine measurement errors, include=FALSE}
measurement_error <-
  read_csv("inst/extdata/measurement_error.csv", show_col_types = FALSE) |>
  group_by(measurement_error) |>
  summarise(values = list(variable)) |>
  deframe()
```

### Missing value

```{r Filter significant correlations of missingness indicators, include=FALSE}
vcf_ms <- paste0(unique(c(vcf$V1, vcf$V2)), "_ms")

correlations_ms <-
  correlation_matrix$data |>
  filter(sig == "Yes") |>
  filter(V1 %in% vcf_ms | V2 %in% vcf_ms) |>
  select(V1, V2)
```

```{r Determine causal direction with missingness indicators, include=FALSE}
ms_directed <- olms_connection(correlations_ms, "_ms$")
```

```{r Determine ol-causal direction with measurement variables, include=FALSE}
m_directed <-
  data.frame(V1 = unique(ms_directed$V2)) |>
  mutate(V2 = str_replace_all(V1, "_ms$", "_m"))

m_directed <-
  data.frame(
    V1 = unique(c(causal_diagram_vcf_data$V1, causal_diagram_vcf_data$V2))
  ) |>
  mutate(V2 = paste0(V1, "_m")) |>
  rbind(m_directed)
```

```{r Determine ol-causal direction of all with missingnesss, include=FALSE}
m_directed_all <-
  data.frame(V1 = colnames(train_data_olms)) |>
  filter(str_detect(V1, "_ms$")) |>
  mutate(V2 = str_replace_all(V1, "_ms$", "_m"))

m_directed_all <-
  m_directed_all |>
  mutate(V1 = str_remove_all(V1, "_ms$")) |>
  rbind(m_directed_all)
```

```{r Obtain CD data with missingness indicators, include=FALSE}
causal_diagram_ms_data <-
  causal_diagram_vcf_data |>
  rbind(m_directed) |>
  rbind(ms_directed) |>
  rbind(m_directed_all) |>
  unique()
```

```{r Create custom function to assess missingness by CD}
assess_ms <- function(data, focal, title = NULL){
  focal <- paste0(focal, c("", "_ms", "_m"))
  
  data <-
    data |>
    filter(V1 %in% focal | V2 %in% focal)
  
  if(nrow(data) == 0){
    NULL
  }else{
    data |>
      correlation_network_plotter(
        layout = "tree"
        , focal = focal[2:3]
        , directed = TRUE
        , title = title
        , curvature = 0.1
      )
  }
}
```

```{r Plot CD with missingness indicators, include=FALSE}
causal_diagram_ms_plot <-
  m_directed_all |>
  filter(str_detect(V1, "_ms$")) |>
  pull(V1) |>
  str_remove_all("_ms$")

causal_diagram_ms_plot <-
  causal_diagram_ms_plot |>
  `names<-`(causal_diagram_ms_plot)

causal_diagram_ms_plot <-
  causal_diagram_ms_plot |>
  imap(~ assess_ms(causal_diagram_ms_data, .x, .y))

causal_diagram_ms_plot <-
  causal_diagram_ms_plot[!sapply(causal_diagram_ms_plot, is.null)]

causal_diagram_ms_plot <-
  ggarrange(
    plotlist = causal_diagram_ms_plot
    , nrow = length(causal_diagram_ms_plot)
  )
```

```{r CD with MS, echo=FALSE, fig.height=250/25.4, fig.width=190/25.4}
causal_diagram_ms_plot
```

```{r Determine mussing-value types, include=FALSE}
missing_value <-
  read_csv("inst/extdata/missing_value.csv", show_col_types = FALSE) |>
  group_by(missing_value) |>
  summarise(values = list(variable)) |>
  deframe()
```

## Causation related to directed cyclic graph (DCG)

```{r Obtain the remaining undirected non-VOIs, include=FALSE}
nonvoi_undirected <-
  correlations |>
  filter(
    V1 %in% c(vcf_directed$V1, vcf_directed$V2)
    & V2 %in% c(vcf_directed$V1, vcf_directed$V2)
  )

nonvoi_undirected <-
  nonvoi_undirected |>
  anti_join(causal_diagram_vcf_data, by = join_by(V1, V2)) |>
  anti_join(
    select(causal_diagram_vcf_data, V1 = V2, V2 = V1)
    , by = join_by(V1, V2)
  ) |>
  mutate(from = NA)
```

```{r Write the undirected non-VOIs to CSV, include=FALSE}
write_csv(nonvoi_undirected, "inst/extdata/nonvoi_undirected.csv")
```

```{r Determine causal direction per pair related to non-VOIs, include=FALSE}
nonvoi_directed <-
  read_csv("inst/extdata/nonvoi_directed.csv", show_col_types = FALSE)
```

```{r Obtain causal diagram data from non-VOIs, include=FALSE}
causal_diagram_nonvoi_data <-
  nonvoi_directed |>
  mutate(to = ifelse(from == 1, 2, 1)) |>
  mutate(
    from = ifelse(from == 1, V1, V2)
    , to = ifelse(to == 1, V1, V2)
  ) |>
  select(V1 = from, V2 = to)
```

```{r Create custom function to find all cycles}
find_all_cycles <- function(edges){
  edges <-
    edges |>
    distinct(from, to) |>
    filter(!is.na(from), !is.na(to))

  nodes <- sort(unique(c(edges$from, edges$to)))
  adj <- split(edges$to, edges$from)

  cycles <- list()

  canon_cycle <- function(path){
    rots <- lapply(
      seq_along(path),
      \(i){
        if(i == 1){
          path
        }else{
          c(path[i:length(path)], path[1:(i - 1)])
        }
      }
    )

    keys <- vapply(
      rots,
      \(x) paste(x, collapse = "->"),
      FUN.VALUE = character(1)
    )

    rots[[order(keys)[1]]]
  }

  dfs <- function(start, current, path, visited){
    nbrs <- adj[[current]]

    if(is.null(nbrs)){
      return(invisible(NULL))
    }

    for(next_node in nbrs){
      if(next_node == start && length(path) >= 2){
        cyc <- canon_cycle(path)
        cycles[[length(cycles) + 1]] <<- cyc
      }else if(!(next_node %in% visited)){
        dfs(
          start = start
          , current = next_node
          , path = c(path, next_node)
          , visited = c(visited, next_node)
        )
      }
    }

    invisible(NULL)
  }

  for(start in nodes){
    dfs(
      start = start
      , current = start
      , path = start
      , visited = start
    )
  }

  if(length(cycles) == 0){
    return(list())
  }

  cycle_keys <- vapply(
    cycles
    , \(x) paste(x, collapse = "->")
    , FUN.VALUE = character(1)
  )

  cycles <- cycles[!duplicated(cycle_keys)]

  lapply(cycles, \(x) c(x, x[1]))
}
```

```{r Create custom function to find all cycles with dataframe output}
find_all_cycles_df <- function(edges){
  cycs <- find_all_cycles(edges)

  map2_dfr(
    cycs
    , seq_along(cycs)
    , \(cyc, id){
      tibble(
        cycle_id = id
        , step = seq_along(cyc)
        , node = cyc
      )
    }
  )
}
```

```{r Find all cycles, include=FALSE}
directed_cyclic_graph <-
  causal_diagram_vcf_data |>
  rbind(causal_diagram_nonvoi_data) |>
  rename(from = V1, to = V2) |>
  find_all_cycles_df()
```

```{r Create custom function to construct causal diagram per cycle}
cycles_to_edge_list <- function(cycles_df){
  cycles_df |>
    dplyr::group_split(cycle_id) |>
    purrr::map(\(df){
      df |>
        dplyr::arrange(step) |>
        dplyr::mutate(
          from = node,
          to = dplyr::lead(node)
        ) |>
        dplyr::filter(!is.na(to)) |>
        dplyr::select(from, to)
    })
}
```

```{r Causal diagram-formatted cycles, include=FALSE}
causal_diagram_dcg_list <-
  cycles_to_edge_list(directed_cyclic_graph) |>
  lapply(rename, V1 = from, V2 = to) |>
  `names<-`(paste0("dcg_", unique(directed_cyclic_graph$cycle_id)))
```

```{r Causal diagram-formatted cycle duplicates, include=FALSE}
causal_diagram_dcg_duplicates <-
  directed_cyclic_graph |>
  group_by(cycle_id) |>
  filter(duplicated(node)) |>
  pull(node) |>
  `names<-`(paste0("dcg_", unique(directed_cyclic_graph$cycle_id)))
```

```{r Plot causal diagram for DCG-related causation, include=FALSE}
causal_diagram_dcg_plot <-
  causal_diagram_dcg_list |>
  imap(
    ~ .x |>
      correlation_network_plotter(
        focal = causal_diagram_dcg_duplicates[[.y]]
        , directed = TRUE, curvature = 0.1
      )
  )

causal_diagram_dcg_plot <-
  ggarrange(
    plotlist = causal_diagram_dcg_plot
    , nrow = length(causal_diagram_dcg_plot)
  )
```

```{r CD for DCGs, echo=FALSE, fig.height=100/25.4, fig.width=190/25.4}
causal_diagram_dcg_plot
```

```{r Caption CD for DCGs, echo=FALSE}
show_figure_legend("causal_diagram_dcg_plot", "Causal diagram showing DGCs separately per duplicate.")
```

## Directional (d)-separation

```{r Create custom function to get descendants of a variable}
get_descendants <- function(var, edges){
  g <-
    igraph::graph_from_data_frame(
      d = edges,
      directed = TRUE
    )
  
  vids <- igraph::V(g)[name == var]
  
  desc <-
    igraph::subcomponent(
      g,
      vids,
      mode = "out"
    )
  
  igraph::V(g)$name[desc] |>
    setdiff(var)
}
```

```{r Identify collider-descendant confounders and mediators, include=FALSE}
coldes_z <-
  path$collider |>
  lapply(
    get_descendants
    , rbind(causal_diagram_vcf_data, causal_diagram_nonvoi_data)
  ) |>
  reduce(c) |>
  intersect(c(path$confounder, path$mediator))
```

```{r Plot causal diagram for coldes cfmed, include=FALSE}
causal_diagram_coldes_z_plot <-
  causal_diagram_vcf_data |>
  rbind(causal_diagram_nonvoi_data) |>
  correlation_network_plotter(
    focal = coldes_z
    , directed = TRUE
    , curvature = 0.1
  )
```

```{r CD for coldes cfmed, echo=FALSE, fig.height=90/25.4, fig.width=190/25.4}
causal_diagram_coldes_z_plot
```

```{r Caption CD for coldes cfmed, echo=FALSE}
show_figure_legend("causal_diagram_coldes_z_plot", "Causal diagram collider-descendant confounders and mediators for VCF related to two VOIs.")
```

```{r Determine a set of variables that provide cond indep, include=FALSE}
z_set <-
  c(path$confounder, path$mediator) |>
  setdiff(coldes_z)
```

```{r Determine formula between VOIs along with Z as covariates, include=FALSE}
covariates <- c(path$causal, z_set)

inference_formula <-
  paste0("outcome", " ~ ", paste0(covariates, collapse = " + ")) |>
  as.formula()
```

# LLM-assisted causal discovery

```{r Variables in VCF with empty definitions, include=FALSE}
vcf_nonvoi_undirected <- rbind(vcf_undirected, nonvoi_undirected)

vcf_var_undefined <-
  data.frame(
    variable = unique(c(vcf_nonvoi_undirected$V1, vcf_nonvoi_undirected$V2))
  ) |>
  mutate(definition = NA)
```

```{r Write variables in VCF with empty definitions, include=FALSE}
write_csv(vcf_var_undefined, "inst/extdata/vcf_var_undefined.csv")
```

Prompt LLM to define variables, e.g.:

=AI("Define this variable. Outcome is preeclampsia, GA is gestational age, suffices of _l/_r mean left and right.", A2)

```{r Determine defition per varibels in VCF, include=FALSE}
vcf_var_defined <-
  read_csv("inst/extdata/vcf_var_defined.csv", show_col_types = FALSE)
```

```{r Create undirected VOIs with confounders for LLM-prompting, include=FALSE}
vcf_prompt_unresponsed <-
  vcf_nonvoi_undirected |>
  select(-from) |>
  left_join(
    rename(vcf_var_defined, V1 = variable, V1_definition = definition)
    , by = join_by(V1)
  ) |>
  left_join(
    rename(vcf_var_defined, V2 = variable, V2_definition = definition)
    , by = join_by(V2)
  ) |>
  mutate(
    prompt =
      paste0(
        "You are a helpful assistant for causal reasoning.\n\n
         Which cause-and-effect relationship is more likely?\n
         A. changing ", V1," causes a change in ", V2,".\n
         B. changing ", V2," causes a change in ", V1,".\n\n
         Definitions:\n
         ", V1,": ", V1_definition,".\n
         ", V2,": ", V2_definition,".\n\n
         Let's work this out in a step by step way to be sure that we have the 
         right answer.\n
         Then provide your final answer within the tags <Answer>A/B</Answer>.
        "
      )
    , response = as.character(NA)
    , answer = as.character(NA)
  )
```

```{r Write the undirected VOIs to CSV for LLM-prompting, include=FALSE}
write_csv(vcf_prompt_unresponsed, "inst/extdata/vcf_prompt_unresponsed.csv")
```

Prompt LLM to extract answer from response, e.g.:

=AI("Extract only the answer.", F2)

```{r Get LLM responses for determining causal direction, include=FALSE}
vcf_prompt_responsed <-
  read_csv("inst/extdata/vcf_prompt_responsed.csv", show_col_types = FALSE)
```

```{r Determine causal direction using LLM responses, include=FALSE}
vcf_directed_llm <-
  vcf_prompt_responsed |>
  mutate(from = ifelse(answer == "A", 1, 2)) |>
  select(V1, V2, from)

nonvoi_directed_llm <-
  vcf_directed_llm |>
  inner_join(select(nonvoi_undirected, -from), by = join_by(V1, V2))

vcf_directed_llm <-
  vcf_directed_llm |>
  inner_join(select(vcf_undirected, -from), by = join_by(V1, V2))
```

```{r Obtain causal diagram data from VCF-LLM, include=FALSE}
causal_diagram_vcf_llm_data <-
  vcf_directed_llm |>
  mutate(to = ifelse(from == 1, 2, 1)) |>
  mutate(
    from = ifelse(from == 1, V1, V2)
    , to = ifelse(to == 1, V1, V2)
  ) |>
  select(V1 = from, V2 = to)
```

```{r Plot causal diagram for VCF-LLM, include=FALSE}
causal_diagram_vcf_llm_plot <-
  causal_diagram_vcf_llm_data |>
  correlation_network_plotter(
    "tree"
    , focal = c("outcome", voi2)
    , directed = TRUE
  )
```

```{r CD for VCF-LLM, echo=FALSE, fig.height=50/25.4, fig.width=190/25.4}
causal_diagram_vcf_llm_plot
```

```{r Caption CD for VCF, echo=FALSE}
show_figure_legend("causal_diagram_vcf_llm_plot", "Causal diagram for VCF based on LLM related to two VOIs without showing edges among non-VOI variables.")
```

```{r Determine path of 2 VOIs based on LLM, include=FALSE}
path_llm <-
  read_csv("inst/extdata/path_llm.csv", show_col_types = FALSE) |>
  group_by(path) |>
  summarise(values = list(variable)) |>
  deframe()
```

```{r Obtain causal diagram data from non-VOIs -LLM, include=FALSE}
causal_diagram_nonvoi_llm_data <-
  nonvoi_directed_llm |>
  mutate(to = ifelse(from == 1, 2, 1)) |>
  mutate(
    from = ifelse(from == 1, V1, V2)
    , to = ifelse(to == 1, V1, V2)
  ) |>
  select(V1 = from, V2 = to)
```

```{r Identify col-des confounders and mediators by LLM, include=FALSE}
coldes_z_llm <-
  path_llm$collider |>
  lapply(
    get_descendants
    , rbind(causal_diagram_vcf_llm_data, causal_diagram_nonvoi_llm_data)
  ) |>
  reduce(c) |>
  intersect(c(path_llm$confounder, path_llm$mediator))
```

```{r Plot causal diagram for coldes cfmed by LLM, include=FALSE}
causal_diagram_coldes_z_llm_plot <-
  causal_diagram_vcf_llm_data |>
  rbind(causal_diagram_nonvoi_llm_data) |>
  correlation_network_plotter(
    focal = coldes_z_llm
    , directed = TRUE
    , curvature = 0.1
  )
```

```{r CD for coldes cfm LLM, echo=FALSE, fig.height=90/25.4, fig.width=190/25.4}
causal_diagram_coldes_z_llm_plot
```

```{r Caption CD for coldes cfmed LLM, echo=FALSE}
show_figure_legend("causal_diagram_coldes_z_llm_plot", "Causal diagram collider-descendant confounders and mediators for VCF based on LLM related to two VOIs.")
```

```{r Determine a set of vars that provide cond indep by LLM, include=FALSE}
z_llm_set <-
  c(path_llm$confounder, path_llm$mediator) |>
  setdiff(coldes_z_llm)
```

```{r Determine formula between VOIs with Z as covariates by LLM, include=FALSE}
covariates_llm <- c(path_llm$causal, z_llm_set)

inference_llm_formula <-
  paste0("outcome", " ~ ", paste0(covariates_llm, collapse = " + ")) |>
  as.formula()
```



















